A thermodynamic view of dusty protoplanetary disks

### 2017-09-12 response.txt

We thank the referee for a prompt report, which helped to improve the
clarity of the manuscript. We have made revisions in bold, and further
explain them below.

Major points:

* The section 2 is interesting as it relates the work done by
  pressure, when not in phase with density, to the growth of
  oscillations. Although I understand that it is a link between the
  pure gas context and the dusty flow case (because the drag force
  acting on the gas can dephase density), I think the authors should
  approach the point later in the paper. In my opinion, it would be
  more adequate to approach it at the end of or after the actual
  section 3 (single fluid description of dusty gas). Then it is
  relevant to make the parallel with the pure gas dynamics. It is also
  a way to explain the analogy that appears in parag. 3.3 "The
  physical reason for this analogy..."

  We appreciate the referee’s suggestion that section 2 might make
  sense if moved.  With respect, our choice for the placement of this
  section is meant for readers who are lack the technical expertise of
  the reviewer. Specifically, our goal is to summarize a main result
  and a key physical insight, which motivates the technical
  derivation of the dusty/adiabatic gas analogy. We added a new
  paragraph at the beginning of section 2 to state its purpose. 
  We hope section 2 motivates further reading and of course any reader
  can revisit  section 2 (section 4.4 refers back to it).

While the referee is certainly correct that the single fluid analogy
needs to made, we have attempted to explain the concept in the
introduction. In any event, we hope the referee will agree to some
authorial prerogative in the order of presentation.

* In section 6 on the numerical results, in particular parag. 6.5 and
  6.6, I wonder which growth rates are obtained with different radial
  mode numbers. It would be an advantage to see other values as in Fig
  8. In 6.5 kx is 30, does the instability change with smaller, larger
  modes? It is worse in 6.6, where kx is very large, 1800. Is it
  relevant to have such a high radial mode, that could barely be
  resolved in non linear numerical simulations? Finally, in a general,
  I think that a more physical summary of the results of the section 6
  should be provided. It is almost limited now to a description of the
  results.

  For section 6.5 we have updated Fig. 14 to include cases with
  different kx. All growth rates remain the same order of
  magnitude. The effect of a radial varation in rhod/rhog is also
  similar between different kx. However, surface modes only appear at
  high kx (as found for pure gas VSI, Barker & Latter, 2017)

  For section 6.6 we also updated Fig. 15 with more kx values. 
  We find to obtain instability with smaller kx, one needs a more
  rapidly-varying dust-to-gas ratio. We do not dwell 
  on this case because we have already shown, using the more general
  analyses in section 4.3 and 5.2

  Note that while the original case with high kx=1800 may not be  
  resolved in global simulations, it can be resolved in local shearing
  boxes. Simulations of SI in fact consider even larger
  kx = O(10^3/h), where h= H/R is the gas aspect-ratio. See, Bai &
  Stone (2010) and Yang & Johansen (2017), for example.

  In section 8 we extend the summary on the results of section
  6. We discuss possible implications of our results in 
  realistic disks with finite stopping times. In particular, we
  suggest the criterion for particles to settle against the VSI.

* The section 7 'Discussion' is for the moment a list of common
  instabilities that could be triggered by the effect of the drag in
  the framework of the one-fluid approximation. Even if it is very
  interesting, it may be too speculative; I would not limit the
  discussion only to this. The authors may first discuss more about
  the limitations/assets of their approach based on the numerical
  results obtained in section 6, which will give arguments to the
  possible numerical implementation of the method (see parag. 7.3). In
  particular, what is the impact of the terminal velocity
  approximation? How much reliable the results and the accuracy of the
  approach are with a finite St number? For example, what do you
  expect for a typical value of 0.01-0.1 (dust or pebble), for which
  it is known that the terminal velocity is inappropriate?

  We have added a new section 7.1 on the application and limitations
  of the one-fluid framework.

  The terminal velocity approximation from the requires ts to be small
  compared to other timescales in the problem. Since 1/Omega is the
  typical `fast' timescale for disks, we generally require ts*Omega <<
  1, but the exact critical value is problem-dependent.

  Fig. 4 showsthat the one-fluid framework can gives accurate
  streaming instability growth rates in dust-rich disks even for
  ts*Omega = 0.1. However, the dust-poor `linB' mode in Table 1 (also
  with ts*Omega=0.1) is not accurately captured by the one-fluid framework. 
  This suggest that the terminal velocity approximation is not the
  only condition required for the one-fluid equations to give accurate
  results. In the case of the streaming instability, it seems that one
  also requires rhod > rhog.

* In the Appendix: parag. D.2.2, the authors check if two assumptions
  done during the calculations are valid (eq D11 and eq D16). But it
  seems that they compute the values with the results obtained after
  using these approximations. Please clarify

  We retitled this section to `Consistency check' since, as the
  referee pointed out, we simply checked that the assumptions made to
  obtain the explicit solutions are self-consistent. The
  approximations are ultimately justified by comparison to numerical
  solutions in Fig. 4.

* Last point: the authors should re read carefully the article, indeed
  some sentences are not completly made agree with their later
  modifications.

  We have proof-read the new manuscript.

Minor points:

* parag 4.4.1 Can you explain briefly in the manuscript how to obtain
  the eq 33. ?

  Additional description (Eq. 33) is now included.

* parag. 5.1.2 Can you clarify the text related to eq 47, one cannot
  see clearly which terms are the second and the third. Maybe one can
  say : "The first term in brackets is usually stabilizing ... "

  This is clarified.

### 2017-09-12 appendix.tex
first draft of revision 1 and response to referee

Eq. \ref{dimless_cubic} now becomes the depressed cubic 
\begin{align}
  \nu^3 - \calP \nu = \calQ, \label{depressed_cubic}
\end{align}
where $\calP$ and $\calQ$ can be read off Eq. \ref{dimless_cubic}. This can be
solved with Vieta's substitution 
\begin{align}
\nu = \mu + \frac{\calP}{3\mu},\label{si_vieta}
\end{align}
then Eq. \ref{depressed_cubic} becomes a quadratic for $\mu^3$,
\begin{align}
\mu^6 - \calQ\mu^3 + \frac{\calP^3}{27} = 0. 
\end{align}
The solution is
\begin{align}
  \mu^3 = \frac{\calQ}{2}\left(1\pm \sqrt{1 -
      \frac{4\calP^3}{27\calQ^2}}\right) 
   = \st \fg (1- 2\fdust)\frac{K_z^2}{K_x}\left\{1 \pm \sqrt{1 -
      \frac{K_z^2}{27K_x^2\left[\st K_x \fg
          \left(1-2\fdust\right)\right]^2}}\right\}.  
\end{align}
At this point we assume the term $\propto K_z^2$ inside the square
root may be neglected. That is, 
\begin{align}
  K_z^2 \ll 27 \left(K_x^2\st\right)^2 \left[\fg
          \left(1-2\fdust\right)\right]^2. \label{max_kz}
%27 K_x^2 \left[\st K_x \fg
%          \left(1-2\fdust\right)\right]^2.
\end{align}
We show in \S\ref{si_dust_rich} that this is readily satisfied. Then
$\mu \simeq \mathcal{Q}^{1/3}$.

@@

\subsubsection{Validity of the approximations}\label{si_dust_rich_approx}
Inserting the solutions for the most unstable mode 
%(Eq. \ref{si_dust_rich_optkx}---\ref{si_dust_rich_grow} with ) 
we find the assumption in Eq. \ref{min_kz} becomes  
\begin{align}
  \fdust \gg \frac{12}{17}\simeq 0.7.
\end{align}
Since $\fdust \leq 1$, this requirement can only be marginally
satisfied. However, we find that this mostly give errors in the real
frequency (see Fig. \ref{si_compare_fig}), while growth rates are
still captured correctly. 
On the other hand, the assumption of Eq. \ref{max_kz} becomes the trivial
inequality \begin{align} 
  1\ll \frac{27}{4}\left(\frac{5}{3}\right)^3\simeq 30,
\end{align}
so Eq. \ref{max_kz} is satisfied, which justifies the approximate solution for 
$\mu$ in Eq. \ref{mu_explicit}.

>>>>

Eq. \ref{dimless_cubic} now becomes the depressed cubic 
\begin{align}
  \nu^3 - \calP \nu = \calQ, \label{depressed_cubic}
\end{align}
where $\calP$ and $\calQ$ can be read off Eq. \ref{dimless_cubic}. This can be
solved with Vieta's substitution 
\begin{align}
\nu = \mu + \frac{\calP}{3\mu},\label{si_vieta}
\end{align}
then Eq. \ref{depressed_cubic} becomes a quadratic for $\mu^3$,
\begin{align}
\mu^6 - \calQ\mu^3 + \frac{\calP^3}{27} = 0. 
\end{align}
The solution is
\begin{align}
  \mu^3 = \frac{\calQ}{2}\left(1\pm \sqrt{1 -
      \frac{4\calP^3}{27\calQ^2}}\right) 
   = \st \fg (1- 2\fdust)\frac{K_z^2}{K_x}\left\{1 \pm \sqrt{1 -
      \frac{K_z^2}{27K_x^2\left[\st K_x \fg
          \left(1-2\fdust\right)\right]^2}}\right\}.  
\end{align}
At this point we assume the term $\propto K_z^2$ inside the square
root may be neglected. That is, 
\begin{align}
  K_z^2 \ll 27 \left(K_x^2\st\right)^2 \left[\fg
          \left(1-2\fdust\right)\right]^2. \label{max_kz}
%27 K_x^2 \left[\st K_x \fg
%          \left(1-2\fdust\right)\right]^2.
\end{align}
We show in \S\ref{si_dust_rich_approx} that this is readily satisfied. Then
$\mu \simeq \mathcal{Q}^{1/3}$.

@@

\subsubsection{Consistency check}\label{si_dust_rich_approx}
{\bf Here we check if the above explicit solutions are
  consistent with the assumptions used to obtain them.  
} 
Inserting the solutions for the most unstable mode 
%(Eq. \ref{si_dust_rich_optkx}---\ref{si_dust_rich_grow} with ) 
{\bf we find Eq. \ref{min_kz} becomes}
\begin{align}
  \fdust \gg \frac{12}{17}\simeq 0.7.
\end{align}
Since $\fdust \leq 1$, this requirement can only be marginally
satisfied. However, we find that this mostly give errors in the real
frequency (see Fig. \ref{si_compare_fig}), while growth rates are
still captured correctly. 
On the other hand, the assumption of Eq. \ref{max_kz} becomes the trivial
inequality \begin{align} 
  1\ll \frac{27}{4}\left(\frac{5}{3}\right)^3\simeq 30,
\end{align}
so Eq. \ref{max_kz} is satisfied, which justifies the approximate solution for 
$\mu$ in Eq. \ref{mu_explicit}.

{\bf Ultimately, the validity of these assumptions are justified
   a posteriori by comparison with the solution to the full equations
   in Fig. \ref{si_compare_fig}. 
}

### 2017-09-12 linear.tex
first draft of revision 1 and response to referee

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
% &1 - \frac{\epsilon
%   h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
%   \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
%   > 0 \label{stability_est2},
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{P}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The second term is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The third term is
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occurred). This means that typical dusty PPDs are stable to
axisymmetric perturbations, even for arbitrarily thin
dust layers.

>>>>

We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$, and that the vertical gas distribution is   
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg2}
becomes 
\begin{align}
  % &1 + \frac{\epsilon}{\left(1+
  %   \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  % \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  % \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
% &1 - \frac{\epsilon
%   h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
%   \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
%   > 0 \label{stability_est2},
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{P}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
for \emph{stability}. The {\bf first term in  brackets} is usually stabilizing in 
PPDs for reasons given in \S\ref{iso_perfect}. The {\bf second term in
brackets} is 
always destabilizing, but is small in smooth, thin disks with radial
gradients $O(1/r)$, $h_\mathrm{g}\ll 1$, 
provided that $H_\epsilon/\Hg$ is not large (e.g. some dust settling
has occurred). This means that typical dusty PPDs are stable to
axisymmetric perturbations, even for arbitrarily thin
dust layers.

### 2017-09-12 model.tex
first draft of revision 1 and response to referee

Noting that only products of perturbations contribute to
$\mathcal{W}$, we find that for periodic oscillations (time dependence 
$e^{\ii\omega t}$ and real $\omega$): 
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
where %the integral is taken over the volume of the
%mixture; and 
 the Lagrangian perturbation $\Delta$ of a variable $X$ is 
$\Delta X = \delta X + \bm{\xi}\cdot\nabla X$ 
and $\bm{\xi}$ is the
Lagrangian displacement ($    \xi_{x,z} =  \ii \dd v_{x,z}/\sigma$).  
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

>>>>

{\bf We consider perturbations such that 
  \begin{align}
 P \to P + \real\left(\Delta P e^{-\ii\sigma t}\right)
 \end{align}
 and similarly for $\rho$. Here, the Lagrangian 
 perturbation $\Delta$ of a variable $X$ is $\Delta X = \delta X + \bm{\xi}\cdot\nabla X$
 and $\bm{\xi}$ is the Lagrangian displacement ($\xi_{x,z} =  \ii\dd
 v_{x,z}/\sigma$). Inserting the above pressure and density fields
 into Eq. \ref{work_def}, and noting that only products of
 perturbations contribute to $\mathcal{W}$ after time-averaging, we
 find for periodic oscillations (time dependence $e^{\ii\omega t}$ and
 real $\omega$) that:  
}
\begin{align}
  \mathcal{W} = - \frac{\omega}{2} \int \imag\left(\Delta P
  \frac{\Delta \rho^*}{\rho}\right)dV,\label{work_real}
\end{align}
Eq. \ref{work_real} show that a phase difference between gas pressure and
the total density leads to work done
($\mathcal{W}\neq0$).

### 2017-09-12 summary.tex
first draft of revision 1 and response to referee

It is important to keep in mind the assumptions 
used to develop our thermodynamic model of dusty gas. 
The terminal velocity approximation, $\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \tstop \nabla P /\rhog$, was employed from the outset. 
%It assumes  
%the dust-gas relative drift $\bm{v}_mathrm{d} - \bm{v}_\mathrm{g} = \tstop \nabla P /\rhog$ 
This is applicable to small particles with 
$\tstop\OmK\ll 1$, but the precise critical value is problem-dependent. For the linear streaming instability,
even $\tstop\OmK \sim 0.1$ can be captured by simplified one-fluid equations 
(Fig. \ref{si_compare_fig}). On the other hand, the simplified equations contain spurious modes that
may overwhelm physical instabilities in certain situations (see \S\ref{spurious_epi}).  %Ultimately, one must check against the full equations.

However, our thermodynamic interpretation of dust-gas drift does \emph{not} in fact depend on the terminal velocity approximation.  
Once the gas equation of state is fixed and the true energy equation deleted, 
the dust continuity equation can be converted to a new effective energy equation. 
%This source term reflects  
%a heat flux proportional to $\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}$, which must generally be solved via a separate 
%evolutionary equation \citep{laibe14}. Nevertheless, this does not change the interpretation of dusty gas as a single ideal fluid subject to cooling.  
As an example, for strictly isothermal dusty gas we obtain   
\begin{align}
\frac{DP}{Dt} = - P\nabla\cdot\bm{v} + c_s^2\nabla\cdot\left[\fdust\left(1-\fdust\right)\rho\left(\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}\right)\right], \label{no_term_vel} 
\end{align}
\citep[see also][]{laibe14}. This equation does not use the terminal velocity approximation (which would give Eq. \ref{eff_energy}), but 
 we see that dust-gas relative drift can be regarded as a heat flux within the mixture. Calculating this heat flux generally requires solving the evolutionary equation for 
$\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}$. Nevertheless, one can still consider the dusty-gas as a single fluid subject to cooling. 
}

@@

%implications for realistic disks
{\bf In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_g$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_g$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}

>>>>

It is important to keep in mind the assumptions 
used to develop our thermodynamic model of dusty gas. 
The terminal velocity approximation, $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g} = \tstop \nabla P /\rhog$, was employed from the
outset. 
%It assumes  
%the dust-gas relative drift $\bm{v}_mathrm{d} - \bm{v}_\mathrm{g} =
%\tstop \nabla P /\rhog$ 
This is applicable to small particles with short stopping times and
strongly coupled to the gas. Generally we require $\tstop$ to be the
shortest timescale in the physical problem. For  
example, to study dust settling, we require $\tstop \ll
t_\mathrm{settle}\sim 1/\tstop\OmK^2$, or 
$\tstop\OmK \ll 1$.

However, the validity of the simplified one-fluid equations may be
problem-dependent, and not just on the terminal velocity 
approximation. For example, Table \ref{si_compare} shows the dust-rich
`linA' mode and dust-poor `linB' mode of the streaming instability
have the same $\tstop$, but the former is accurately captured by the
simplified equations, while the latter is not.  This suggests that for
the streaming instability the simplified equations is better suited
for $\rhod/\rhog>1$. The simplified equations may also contain spurious
modes in certain situations (see \S\ref{spurious_epi}).

%Ultimately,
                                %one must 
                                %c heck against the full equations.

Note that our thermodynamic interpretation of dust-gas drift does
\emph{not} in fact depend on the terminal velocity approximation.   
Once the gas equation of state is fixed and the true energy equation deleted, 
the dust continuity equation can be converted to a new effective energy equation. 
As an example, for strictly isothermal dusty gas we obtain   
\begin{align}
\frac{DP}{Dt} = - P\nabla\cdot\bm{v} +
c_s^2\nabla\cdot\left[\fdust\left(1-\fdust\right)\rho\left(\bm{v}_\mathrm{d}
    - \bm{v}_\mathrm{g}\right)\right]. \label{no_term_vel}  
\end{align}
This equation does not use
the terminal velocity approximation (which gives
Eq. \ref{eff_energy}); evaluating the right-hand-side generally 
requires solving an evolutionary equation for $\bm{v}_\mathrm{d} - 
\bm{v}_\mathrm{g}$ \citep{laibe14}. Nevertheless,
Eq. \ref{no_term_vel} shows that dust-gas relative drift can be
interpreted as a heat flux within the dust-gas mixture. Thus, even 
without the terminal velocity approximation we can 
regard the dusty-gas as a single ideal fluid subject to cooling.   
}

@@

%implications for realistic disks
{\bf In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_\mathrm{g}$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_g$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}
